---
title: 'Reproducible Research: Peer Assessment 1'
output:
  html_document:
    keep_md: yes
  pdf_document: default
  word_document: default
---


## Loading and preprocessing the data

To load the dataset we first need to extract to source zip file. Note that we only perfom this step if the activity.csv file doesn't exist yet:

```{r}
if(!file.exists("activity.csv"))
{
  print("Extracting activity dataset...")
  unzip("activity.zip");
}
```

Load we can load the dataset and preprocess it. First we will format the date column into POSIXct values:

```{r}
data <- read.csv("activity.csv")
data$date <- as.POSIXct(data$date)
head(data,10)
```

## What is mean total number of steps taken per day?

To compute the mean total number of steps taken per day we first need to compute the total number of steps for each day.
This can be achieved with the dplyr and summarise functions:

```{r}
library(plyr)
total_steps_per_day <- ddply(data,"date",summarize,total_steps=sum(steps))
head(total_steps_per_day,10)
```

Let's now display an histogram of the total number of steps per day:

```{r}
hist(total_steps_per_day$total_steps,breaks=10, main="Number of steps per day")
```

Finaly we can compute the mean of the total number of steps simply ignoring the NA values, as well as the median value:

```{r}
mean_tot_steps <- mean(total_steps_per_day$total_steps, na.rm = T)
mean_tot_steps
med <- median(total_steps_per_day$total_steps, na.rm = T)
med
```

The mean of the total number of steps per day is of about `r format(floor(mean_tot_steps+.5),scientific=F)` steps, whereas the median value is around `r format(floor(med+.5),scientific=F)`

## What is the average daily activity pattern?

To answer this question we start with building the mean number of steps for each 5 minutes chunk:

```{r}

mean_pat <- ddply(data,"interval",summarize,mean_steps=mean(steps, na.rm = T))
head(mean_pat,10)
```

Now we can plot this mean steps per day partern, and subsequently compute the maximum mean step value

```{r}
# Compute the max mean step value and the corresponding interval:
plot(x=mean_pat$interval,y=mean_pat$mean_steps, type='l',main="Mean steps pattern",
     xlab="5-Minute intervals", ylab="Mean steps value")

maxi <- max(mean_pat$mean_steps)
maxint <- mean_pat$interval[mean_pat$mean_steps == maxi]

# compute the corresponding day time:
hours <- floor(maxint/60.0)
mins <- maxint - hours*60

daytime <- sprintf("%02d:%02d", hours,mins)
```


As reported in the previous plot the maximum number of mean steps is of `r maxi`, and it is detected here in the interval `r maxint`, which correspond to the day time of `r daytime` (assuming that the interval 0 corresponds to time 00:00).


## Imputing missing values

Let's now compute the number of rows where we have NA values:

```{r}

numna <- sum(is.na(data$steps))
numna

# compute the proportion of NA:
prop <- sprintf("%.2f",100.0*numna/dim(data)[1])
```

In this dataset we have `r numna` NA rows, which represents `r prop`% of the available data.


Lets now fill those missing data in the dataset: for each row where we have an NA steps value, we will simply replace this NA value with the mean steps value corresponding to that 5 minutes interval:

```{r}

# create a copy of the dataset:
filled_data <- data

# Fill the missing data with a simple for loop:
steps <- data$steps
for(i in seq_along(steps))
{
  if(is.na(steps[i]))
  {
    steps[i] <- mean_pat$mean_steps[mean_pat$interval == data$interval[i]]
  }
}

filled_data$steps <- steps
head(filled_data,10)
```

Now we build a new histogram from the filled dataset:

```{r}
tot_steps <- ddply(filled_data,"date",summarize,total_steps=sum(steps))
hist(tot_steps$total_steps,breaks=10, main="Number of steps per day (filled data)")
tot_mean <- mean(tot_steps$total_steps)
tot_mean
tot_med <- median(tot_steps$total_steps)
tot_med
```

We observe almost the same values for the mean and median values when using this filled dataset. So filling the missing values seem to have basically no effect in this case.


## Are there differences in activity patterns between weekdays and weekends?

First we prepare a factor to separate the week days from the weed end days:

```{r}
days <- weekdays(data$date)
weekends <- days=="Saturday" | days=="Sunday"

# add this vector as a factor in the filled dataset:
filled_data$weekends <- factor(weekends)

# rename the levels:
levels(filled_data$weekends) <- c("weekday","weekend")
```


Now we can compute the mean steps for each 5 minutes interval and for each week day type using again the summarize function:

```{r}
mean_steps_week <- ddply(filled_data,c("interval","weekends"),summarize,mean_steps=mean(steps))
head(mean_steps_week,10)
```

Lets now draw a plot to compare those mean steps patterns:

```{r}
library(lattice)
with(mean_steps_week,xyplot(mean_steps~interval|factor(weekends),
       type='l',layout=c(1,2),
       xlab='Interval',ylab='Number of Steps'))
```

As we can see from the previous graph the patterns during week days and week ends are a bit different: on the whole it seems the number of steps are more evenly distributed on the complete days during week ends.

